Persistence in functional topology: q-tameness, local-connectivity and a correction to a theorem of Morse

### 2021-03-31 connectivity.tex
Changes before meeting

\anibal{Use a different example, $\H$ not yet defined.}
\begin{ex}
	For any $d$, the $d$-dimensional Hawaiian earring is not $\HLC$. However, if we consider $\H$ as a subspace of $\mathbb{H}^{d'}$ via the obvious embedding for $d < d'$, then $\H$ is $\HLC$ in $\mathbb{H}^{d'}$.
\end{ex}

Local connectedness assumptions have a long history of being used to ensure that different homology theories agree and to ensure that homology is finite dimensional, see \cite{MR0007094} for early examples. For us, the following is relevant.

@@

\begin{cor}\label{cor:cech_sing_persistent_iso}
	Let $M$ be a metric space and $f\colon M\to\mathbb{R}$ a function such that $f_{\leq c}$ is a paracompact $HLC^\infty$-space for each $c\in\mathbb{R}$. Then $\CH_d(f_{\leq\bullet})\cong H_d(f_{\leq\bullet})$ for any $d$.
\end{cor}

@@

\begin{prop}\label{prop:fin_dim_sing_hom}
	Let $X$ be compact Hausdorff and $HLC^{\infty}$. Then $H_d(X;\mathbb{F})$ is finite-dimensional for all $d$ and any field $\mathbb{F}$.
\end{prop}

@@

\begin{cor}
	Let $M$ be a topological space and $f\colon M\to\mathbb{R}$ a function such that $f_{\leq c}$ is a compact Hausdorff $HLC^\infty$-space for each $c\in\mathbb{R}$. Then $H_d(f_{\leq\bullet})$ is p.f.d. for any $d$.
\end{cor}

>>>>

\anibal{Use a different example, $\H$ not yet defined.}
\begin{ex}
	The $d$-dimensional Hawaiian earring $\H$ is not $\HLC$. However, if we consider $\H$ as a subspace of $\mathbb{H}^{d'}$ via the obvious embedding for $d < d'$, then $\H$ is $\HLC$ in $\mathbb{H}^{d'}$.
\end{ex}

Local connectivity assumptions have a long history of being used to ensure that different homology theories agree and to ensure that homology is finite dimensional, see \cite{MR0007094} for early examples. For us, the following is relevant.

@@

\begin{cor}\label{cor:cech_sing_persistent_iso}
	Let $M$ be a metric space and $f \colon M \to \mathbb{R}$ a function such that $M_{\leq c}$ is a paracompact $\HLC$-space for each $c \in \mathbb{R}$.
	Then $\CH_n(M_{\leq \bullet}) \cong H_n(M_{\leq\bullet})$ for any integer $n$.
\end{cor}

@@

\anibal{Does this needs field coefficients? (}
\begin{prop} \label{prop:fin_dim_sing_hom}
	Let $X$ be compact Hausdorff and $\HLC$.
	Then $H_n(X; \mathbb{F})$ is finite-dimensional for all $n$ and any field $\mathbb{F}$.
\end{prop}

@@

\begin{cor}
	Let $X$ be a topological space and $f \colon X \to \R$ a function such that $f_{\leq c}$ is a compact Hausdorff $\HLC$-space for each $c \in \R$.
	Then $H_d(f_{\leq\bullet})$ is p.f.d. for any $d$.
\end{cor}

### 2021-03-31 introduction.tex
Changes before meeting

\begin{defi}
	Let $M$ be a metric space and $f\colon M\to\mathbb{R}$ a function. The space $M$ is called \emph{locally-$f$-connected} if for each $p\in M$, $e>0$ and $d\in\mathbb{N}_0$ there exists $\delta\in(0,e)$ such that the map 
	\[
	\tilde{H}_d(f_{\leq f(p)+\delta}\cap B_{\delta}(p)\hookrightarrow f_{\leq f(p)+e}\cap B_e(p);\mathbb{Z})
	\]
	is 0.
\end{defi}

Here, $\tilde{H}_{d}$ denotes reduced $d$-dimensional singular homology, $f_{\leq t}$ denotes the sublevel set of $f$ at $t\in\mathbb{R}$, and $B_{\delta}(p)$ denotes the open ball with radius $\delta$ around the point $p$. Morse then states the claim below as \cite[Theorem 6.3]{Morse.1940}.

\begin{claim}
	Let $M$ be a metric space and $f\colon M\to\mathbb{R}$ a function such that $M$ is locally-$f$-connected and all sublevel sets are compact. Then $\CH_d(f_{\leq\bullet};\mathbb{F})$ is q-tame for any $d$ and any field $\mathbb{F}$.
\end{claim}

Here, $\CH_{d}$ denotes $d$-dimensional Vietoris or \v{C}ech homology and $f_{\leq\bullet}$ denotes the sublevel set filtration of $f$. He does not prove the statement in the above reference, but cites \cite[Theorem 6.1]{Morse.1938}. There, Morse presents a supposed proof of the claim by constructing finite bases for the sets $\im \CH_d(f_{\leq s}\hookrightarrow f_{\leq t};\mathbb{F})$. As it turns out, the proof is not correct: We present a counterexample to the claim.

\begin{thm}\label{thm:counterexample}
	For any $d$, there exists a funtion $F\colon \mathbb{H}^{d}\to\mathbb{R}$ on the $d$-dimensional Hawaiian earring $\mathbb{H}^{d}$ such that $\mathbb{H}^{d}$ is locally-$F$-connected, all sublevel sets are compact and $\CH_d(F_{\leq\bullet};\mathbb{F})$ is not q-tame for any field $\mathbb{F}$.
\end{thm}

Recall that the $d$-dimensional Hawaiian earring is defined as the subspace
\[
\mathbb{H}^{d}=\bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\mathbb{R}^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}
\]
of $d+1$-dimensional Euclidean space.

@@

\begin{defi} 
	Let $M$ be a metric space and $f\colon M\to\mathbb{R}$ a function. The space $M$ is called \emph{locally-$f$-connected} if for each $p \in M$, $e > 0$ and $d \in \mathbb{N}_0$ there exists $\delta \in (0,e)$ such that for all $c \geq f(p)$ the map 
	\[
	\tilde{H}_d(f_{\leq c+\delta} \cap B_{\delta}(p) \hookrightarrow f_{\leq c+e} \cap B_e(p); \mathbb{Z})
	\]
	is 0.
\end{defi}

@@

\begin{claim}\label{con:q-tame_pers_cech_hom}
	Let $M$ be a metric space and $f\colon M\to\mathbb{R}$ a function such that $M$ is strongly locally-$f$-connected and all sublevel sets are compact. Then $\CH_d(f_{\leq\bullet};\mathbb{F})$ is q-tame for any $d$ and any field $\mathbb{F}$.
\end{claim}

While we are currently unable to present a proof, we conjecture the claim to be true.

\todo{what does the story look like for cohomology?}

\subsection*{Notation and Conventions}
For the rest of the paper, we let $\mathbb{F}$ be a field and whenever we write homology without explicit mention of coefficients, we mean homology with coefficients in $\mathbb{F}$.

>>>>

\begin{defi}
	Let $M$ be a metric space and $f \colon M \to \R$ a function. The space $M$ is called \emph{locally-$f$-connected} if for each $p\in M$, $e>0$ and $d\in\mathbb{N}_0$ there exists $\delta\in(0,e)$ such that the map 
	\[
	\tilde{H}_d(f_{\leq f(p)+\delta}\cap B_{\delta}(p)\hookrightarrow f_{\leq f(p)+e}\cap B_e(p);\mathbb{Z})
	\]
	is 0.
\end{defi}

Here, $\tilde{H}_{d}$ denotes reduced $d$-dimensional singular homology, $f_{\leq t}$ denotes the sublevel set of $f$ at $t\in\R$, and $B_{\delta}(p)$ denotes the open ball with radius $\delta$ around the point $p$. Morse then states the claim below as \cite[Theorem 6.3]{Morse.1940}.

\begin{claim}
	Let $M$ be a metric space and $f\colon M\to\R$ a function such that $M$ is locally-$f$-connected and all sublevel sets are compact. Then $\CH_d(f_{\leq\bullet};\mathbb{F})$ is q-tame for any $d$ and any field $\mathbb{F}$.
\end{claim}

Here, $\CH_{d}$ denotes $d$-dimensional Vietoris or \v{C}ech homology and $f_{\leq\bullet}$ denotes the sublevel set filtration of $f$. He does not prove the statement in the above reference, but cites \cite[Theorem 6.1]{Morse.1938}. There, Morse presents a supposed proof of the claim by constructing finite bases for the sets $\im \CH_d(f_{\leq s}\hookrightarrow f_{\leq t};\mathbb{F})$. As it turns out, the proof is not correct: We use the Hawaiian earring to construct a counterexample to this claim in Section \ref{sec:counterexample}.

@@

\begin{defi} 
	Let $M$ be a metric space and $f\colon M\to\R$ a function. The space $M$ is called \emph{locally-$f$-connected} if for each $p \in M$, $e > 0$ and $d \in \mathbb{N}_0$ there exists $\delta \in (0,e)$ such that for all $c \geq f(p)$ the map 
	\[
	\tilde{H}_d(f_{\leq c+\delta} \cap B_{\delta}(p) \hookrightarrow f_{\leq c+e} \cap B_e(p); \mathbb{Z})
	\]
	is 0.
\end{defi}

@@

\begin{claim}
	Let $M$ be a metric space and $f \colon M \to \R$ a function such that $M$ is strongly locally-$f$-connected and all sublevel sets are compact.
	Then $\CH_d(f_{\leq\bullet};\mathbb{F})$ is q-tame for any $d$ and any field $\mathbb{F}$.
\end{claim}

We devote Section \ref{sec:theorem} to its proof.

### 2021-03-31 preliminaries.tex
Changes before meeting

\begin{itemize}
	\item metric space
	\item upper semi-continuous function
	\item Persistence module
	\item q-tameness
	\item Barcode
	\item bottleneck
	\item stability
\end{itemize}

@@

>>>>

\subsection*{Notation and Conventions}
For the rest of the paper, we let $\mathbb{F}$ be a field and whenever we write homology without explicit mention of coefficients, we mean homology with coefficients in $\mathbb{F}$.

\subsection{Sublevel set filtrations}

Treat in parallel the case of spaces $X$ and the case of sublevel filtrations $(X, f)$. Leads nicely to different connectivities.
\begin{itemize}
	\item metric space
	\item upper semi-continuous function
	\item sublevel set filtrations
\end{itemize}

@@

\subsection{Persistence homology}

\begin{itemize}
	\item Persistence module
	\item q-tameness
	\item Barcode
	\item bottleneck
	\item stability
\end{itemize}

### 2021-03-31 theorem.tex
Changes before meeting

\documentclass{amsart}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{tikz-cd}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\HS}{\mathrm{HS}}
\newcommand{\dprime}{{\prime\prime}}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\interior}{int}

\setlength{\textwidth}{\paperwidth}
\addtolength{\textwidth}{-1.4in}
\setlength{\textheight}{\paperheight}
\addtolength{\textheight}{-1.8in}
\calclayout

\newcommand{\p}{\mathrm{p}}
\newcommand{\q}{\mathrm{q}}

\begin{document}

\begin{definition}
	For $n \in \Z$ a continuous map is said to be \textit{$n$-homologically small} ---denoted $\HS_n$--- if the image of the map induced by $H_n(-)$ is finitely generated, and we say it is \textit{homologically small}, denoted $\HS$, if it is $\HS_n$ for every~$n$.
\end{definition}

\begin{definition}
	Let $X$ be a space and $f \colon X \to \R$ a function. For any $t \in \R$, the \textit{$t$-sublevel set of $f$} is defined by 
	\begin{equation*}
	X_{\leq t} = f^{-1}((-\infty, t]).
	\end{equation*}
\end{definition}

\begin{definition}
	Let $M$ be a metric space. For any $p \in M$ and $\epsilon \geq 0$ denote
	\begin{equation*}
	B_\epsilon(p) = \{q \in M\ |\ d(p,q) < \epsilon\}, \qquad
	\overline B_\epsilon(p) = \{q \in M\ |\ d(p,q) \leq \epsilon\}.
	\end{equation*}
\end{definition}

\begin{definition}
	Let $M$ be a metric space and $f \colon M \to \R$ a function.
	The space $M$ is said to be \textit{strongly locally-$f$-connected} if for each $p \in M$, $e > 0$, and $n \in \N$ there exists $\delta \in (0, e)$ such that for every $c \geq 0$ the image of the map induced by $H_n(-)$ on $B_\delta(p) \cap M_{\leq f(p)+\delta+c} \to B_e(p) \cap M_{\leq f(p)+e+c}$ is $0$.
\end{definition}

\begin{theorem} \label{t:strong local connectenss implies q-tameness}
	Let $M$ be a metric space and $f \colon M \to \R$ a function. If $M$ is strongly locally-$f$-connected and all level sets are compact, then, for any $s < t$ the inclusion $M_s \to M_t$ is $\HS$.
\end{theorem}

\begin{lemma} \label{l:commutative algebra}
	Given a commutative diagram of modules over a principal ideal domain
	\begin{equation*}
	\begin{tikzcd}
	A_{1,1} \arrow[r] & A_{1,2} & \\
	A_{2,1} \arrow[r] \arrow[u] & A_{2,2} \arrow[r] \arrow[u] & A_{2,3} \\
	& A_{3,2} \arrow[r] \arrow[u] & A_{3,3} \arrow[u]
	\end{tikzcd}
	\end{equation*}
	where the middle row is exact and both $A_{2,1} \to A_{1,1}$ and $A_{3,3} \to A_{2,3}$ have finitely generated images, then $A_{3,2} \to A_{1,2}$ does as well.
\end{lemma}

\begin{lemma} \label{l:neighborhood third}
	Let $M$ be a metric space. For any compact subset $K$ and open set $U$ with $K \subseteq U$ there exists a compact set $K^\prime$ such that
	\begin{equation*}
	K \subseteq \interior(K^\prime) \subseteq K^\prime \subseteq U.
	\end{equation*}
\end{lemma}

@@

\begin{lemma} \label{l:key lemma for q-tameness}
	Let $M$ and $f$ be as in Theorem \ref{t:strong local connectenss implies q-tameness}.
	Consider sets $K, L \subseteq M$ with $K$ compact and $K \subseteq \interior(L)$. For any $s < t$ there is $\delta \in (0,\, t-s)$ such that $K \cap M_{s+\delta} \to L \cap M_{t}$ is $\HS$.
\end{lemma}

@@

\end{document}

>>>>

\section{Theorem} \label{sec:theorem}

\begin{thm} \label{t:strong local connectenss implies q-tameness}
	Let $M$ be a metric space and $f \colon M \to \R$ a function. If $M$ is strongly locally-$f$-connected and all level sets are compact, then, for any $s < t$ the inclusion $M_s \to M_t$ is $\HS$.
\end{thm}

\begin{lem} \label{l:commutative algebra}
	Given a commutative diagram of modules over a principal ideal domain
	\begin{equation*}
	\begin{tikzcd}
	A_{1,1} \arrow[r] & A_{1,2} & \\
	A_{2,1} \arrow[r] \arrow[u] & A_{2,2} \arrow[r] \arrow[u] & A_{2,3} \\
	& A_{3,2} \arrow[r] \arrow[u] & A_{3,3} \arrow[u]
	\end{tikzcd}
	\end{equation*}
	where the middle row is exact and both $A_{2,1} \to A_{1,1}$ and $A_{3,3} \to A_{2,3}$ have finitely generated images, then $A_{3,2} \to A_{1,2}$ does as well.
\end{lem}

\begin{lem} \label{l:neighborhood third}
	Let $M$ be a metric space. For any compact subset $K$ and open set $U$ with $K \subseteq U$ there exists a compact set $K^\prime$ such that
	\begin{equation*}
	K \subseteq \interior(K^\prime) \subseteq K^\prime \subseteq U.
	\end{equation*}
\end{lem}

@@

\begin{lem} \label{l:key lemma for q-tameness}
	Let $M$ and $f$ be as in Theorem \ref{t:strong local connectenss implies q-tameness}.
	Consider sets $K, L \subseteq M$ with $K$ compact and $K \subseteq \interior(L)$. For any $s < t$ there is $\delta \in (0,\, t-s)$ such that $K \cap M_{s+\delta} \to L \cap M_{t}$ is $\HS$.
\end{lem}

@@
